---
title: "Cyclistic_Analysis"
author: "Joanne Fil"
date: "8/13/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Case Study 1: Cyclistic Bike-Share Analysis

We need to understand how annual members and casual riders use Cyclistic bikes differently.

### Install packages and set directory

```{r cars}
library(tidyverse)  #helps wrangle data
library(lubridate)  #helps wrangle date attributes
library(ggplot2)  #helps visualize data
library(plyr)
library(dplyr)
```

## Step 1: Collect data

Upload data

```{r pressure, echo=FALSE}
April_2020 <- read.csv("202004-divvy-tripdata.csv")
May_2020 <-  read.csv("202005-divvy-tripdata.csv")
June_2020 <-  read.csv("202006-divvy-tripdata.csv")
July_2020 <-  read.csv("202007-divvy-tripdata.csv")
August_2020 <-  read.csv("202008-divvy-tripdata.csv")
September_2020 <-  read.csv("202009-divvy-tripdata.csv")
October_2020 <-  read.csv("202010-divvy-tripdata.csv")
November_2020 <-  read.csv("202011-divvy-tripdata.csv")
December_2020 <-  read.csv("202012-divvy-tripdata.csv")
January_2021 <-  read.csv("202101-divvy-tripdata.csv")
February_2021 <-  read.csv("202102-divvy-tripdata.csv")
March_2021 <-  read.csv("202103-divvy-tripdata.csv")
```

## Step 2: Wrangle data and combine into a single file

We want to eventually join the files together, so must first compare the column names to ensure they are the same.


```{r}
colnames(April_2020)
colnames(May_2020)
colnames(June_2020)
colnames(July_2020)
colnames(August_2020)
colnames(September_2020)
colnames(October_2020)
colnames(November_2020)
colnames(December_2020)
colnames(January_2021)
colnames(February_2021)
colnames(March_2021)
```

Column names were the same. If they were not, we would have renames them using 'rename'.

Now we examine the structure of the dataframes to ensure they are the same. We can do this individually:

```{r}
str(April_2020)
str(May_2020)
str(June_2020)
str(July_2020)
str(August_2020)
str(September_2020)
str(October_2020)
str(November_2020)
str(December_2020)
str(January_2021)
str(February_2021)
str(March_2021)
```

or at the same time using a loop:

```{r}
dataImport_list <- list(April_2020, May_2020, June_2020, July_2020, August_2020, September_2020,
                        October_2020, November_2020, December_2020, January_2021,
                        February_2021, March_2021) # combine the data sets into a list
for(i in dataImport_list) {
  str(i)
}
```

or view the structure in a condensed version:

```{r}
janitor::compare_df_cols(dataImport_list)
```
Some data frames have 'start_station_id' and 'end_station_id' as 'int' when they should be 'chr'. Let's convert them to 'chr' so that they can stack correctly. 

```{r}
April_2020 <- mutate(April_2020, start_station_id = as.character(start_station_id), end_station_id = as.character(end_station_id))
May_2020 <- mutate(May_2020, start_station_id = as.character(start_station_id), end_station_id = as.character(end_station_id))
June_2020 <- mutate(June_2020, start_station_id = as.character(start_station_id), end_station_id = as.character(end_station_id))
July_2020 <- mutate(July_2020, start_station_id = as.character(start_station_id), end_station_id = as.character(end_station_id))
August_2020 <- mutate(August_2020, start_station_id = as.character(start_station_id), end_station_id = as.character(end_station_id))
September_2020 <- mutate(September_2020, start_station_id = as.character(start_station_id), end_station_id = as.character(end_station_id))
October_2020 <- mutate(October_2020, start_station_id = as.character(start_station_id), end_station_id = as.character(end_station_id))
November_2020 <- mutate(November_2020, start_station_id = as.character(start_station_id), end_station_id = as.character(end_station_id))
```

Now let's stack the individual data frames into one large data frame

```{r}
all_trips <- bind_rows(April_2020, May_2020, June_2020, July_2020, August_2020, September_2020,
                       October_2020, November_2020, December_2020, January_2021,
                       February_2021, March_2021)
```

## Step 3: Clean up and add data to prepare for analysis

Let's inspect the new table that has been created

```{r}
colnames(all_trips)  #List of column names
```

```{r}
nrow(all_trips)  #How many rows are in data frame?
```

```{r}
dim(all_trips)  #Dimensions of the data frame?
```

```{r}
head(all_trips)  #See the first 6 rows of data frame
```

```{r}
tail(all_trips)#See the last 6 rows of data frame
```

```{r}
str(all_trips)  #See list of columns and data types (numeric, character, etc)
```

Now we should make sure that there are only two categories of customers by looking at the number of unique values in the 'member_casual' column

```{r}
unique(all_trips$member_casual) 
```

or by looking at the number of subjects within each category in the 'member_casual' column

```{r}
table(all_trips$member_casual)
```

*Note* If there was more than two categories of customers (for example had 'subscriber' instead of 'member'), we would reassign the values by:
alltrips <- all_trips %>% 
mutate(member_casual = recode(member_casual,"subscriber" - "member"))

Now we want to split up day, month, and year of the ride. Here is a good website explaining the various types of ways you can get date information: https://www.statmethods.net/input/dates.html 

```{r}
all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

Let's calculate the length of the ride (in seconds), but first we should inspect the structure of the columns to make sure the length of ride is numeric and not a factor

```{r}
is.factor(all_trips$ride_length)
```

The output was 'false'FALSE' meaning it was not a factor

```{r}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
```

Now let's make sure there are no negative values for the ride length (when bikes were taken out of docks and checked for quality)

```{r}
count(all_trips$ride_length < 0)
```

There are some negative values (they were presented as 'TRUE') so we will create a new data frame to not include the ones that are negative

```{r}
all_trips_v2 <- all_trips[!(all_trips$ride_length < 0),]
```

## Step 4: Conduct descriptive analysis

Let's do descriptive analysis on ride_length (in seconds)

Mean

```{r}
mean(all_trips_v2$ride_length) #straight average (total ride length / rides) 
```

Median

```{r}
median(all_trips_v2$ride_length) #midpoint number in the ascending array of ride lengths
```

Maximum

```{r}
max(all_trips_v2$ride_length) #longest ride
```

Minimum

```{r}
min(all_trips_v2$ride_length) #shortest ride
```

Let's compare members and casual users using 'aggregate' (similar to pivot table)

Mean

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
```

Median

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median) 
```

Maximum

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max) 
```

Minimum

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min) 
```

Now let's observe the average ride time by each day for members vs casual users

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

The days of the week are out of order. Let's fix that.

```{r}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

Now, let's run the average ride time by each day for members vs casual users

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

Let's analyze ridership data by type and weekday

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  #groups by usertype and weekday
  dplyr::summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% # calculates the number of rides and average duration and calculates the average duration
  arrange(member_casual, weekday)			# sorts
```

 Let's visualize the number of rides by rider type
 
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  dplyr::summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```
 
Let's create a visualization for average duration

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  dplyr::summarise(number_of_rides = n()
            ,average_duration = mean(ride_length/60)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") + xlab("Weekday") + ylab("Average Duration (minutes)")
```

Let's create a visualization for number of rides each month 

```{r}
ggplot(all_trips_v2) +
  geom_bar(mapping = aes(x = month, fill = member_casual), position = "dodge", width = 0.6) + 
  xlab("Month") + ylab("Number of Rides")
```

Let's create a visualization for average duration for each month 

```{r}
all_trips_v2 %>% 
  group_by(member_casual, month) %>% 
  dplyr::summarise(number_of_rides = n()
                   ,average_duration = mean(ride_length/60)) %>% 
  arrange(member_casual, month)  %>% 
  ggplot(aes(x = month, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge", width = 0.65) + xlab("Month") + ylab("Average Duration (minutes)")
```

Look if type of bike has influence

```{r}
ggplot(all_trips_v2) +
  geom_bar(mapping = aes(x = rideable_type, fill = member_casual), position = "dodge", width = 0.6) + 
  xlab("Type of Bike") + ylab("Number of Rides")
```

## Step 5: Export summary files for further analyses
Create a csv file that we will visualize in Tableau

First create data set that only has member type, day of the week, and average ride length 

```{r}
counts <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
write.csv(counts, file = 'avg_ride_length.csv')
```

Then create data set that has ride type, member type, the latitude and longitude of ride start and end, day of week, and ride length

```{r}
condensed <- all_trips_v2 %>% 
  select(rideable_type, start_lat, start_lng, end_lat, end_lng, member_casual, day_of_week, ride_length)

write.csv(condensed, file = ('Trips_condensed.csv'))
```












